---
title: "Fastq format sequence files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fastq format sequence files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Preamble

This `fastq_processing` vignette re-introduces some of the content from the
`BasicQC` tutorial concept but derives the primary information from the FASTQ
file for users who have not maintained their sequencing_summary file. This is
intended to facilitate the development of workflows and reports that are
decoupled from the requirement for the `sequencing_summary` file.


## Getting started with the canned dataset

The `floundeR` package is distributed with a collection of canned datasets.
These include an example `FASTQ` file that has been gzip compressed and contains
a somewhat lacklustre historical dataset that is interesting only in its
compactness.

```{r setup}
library(floundeR)

canonical_fastq <- flnDr("example.fastq.gz")
fastq <- Fastq$new(canonical_fastq)

print(fastq)

fastq$as_tibble()
```

So what have we done here? We have identified the packaged fastq file and we
have used this file to instantiate the `Fastq` object - this can be displayed
using the `print()` command and we can have a quick look at the data that has
been extracted using the `$as_tibble()` function that is exported by the 
package.

```{r simple_sequenceset}

fastq %>% to_sequencing_set()

```

The `SequencingSet` in turn has a collection of methods that can be used to
structure and visualise the data. The first that we'll have a look at is the
`$enumerate` method that returns an `Angenieux` object for data visualisation.

```{r sequencingset_enumerate, out.width="75%"}
knitr::include_graphics(
  fastq$sequencingset$enumerate$to_file("figure_5.png")$plot())
```

There are a plethora of ways through which the `Angenieux` object can be used
to style, colour and manipulate the graph - please do have a look at the methods
documentation.

The `SequencingSet` object can also be used to access simple but primitive
summary statistics such as **`mean`** sequence length, **`N50`** length etc

```{r sequencingset_stats}
fastq$sequencingset$N50
fastq$sequencingset$mean
```


### Review sequence length distributions

The distribution of sequence lengths is an important metric that is impacted by
choice of library preparation, starting DNA isolation etc. A plot of length
distributions is prepared from the same `SequencingSet` object that we
reviewed in the previous section.

```{r sequencingset_readlengths, out.width="75%"}
knitr::include_graphics(
  fastq$sequencingset$read_length_bins(bins=35, outliers=0.001)$
    to_file("figure_6.png")$
    plot(style="stacked"))

```


### Review quality distributions

The distribution of sequence lengths is an important metric that is impacted by
choice of library preparation, starting DNA isolation etc. A plot of length
distributions is prepared from the same `SequencingSet` object that we
reviewed in the previous section.

```{r sequencingset_qualitybins, out.width="75%"}
knitr::include_graphics(
  fastq$sequencingset$quality_bins(bins=100)$
    to_file("figure_7.png")$
    plot(style="stacked"))

```


## Exploring a more impressive sequence collection

The `Guppy` basecalling software converts the Nanopore format FAST5 raw sequence
files into the FASTQ files that we have reviewed in the previous section. The
FASTQ entries prepared by `Guppy` contain additional information in their
**header** fields. These additional information contain metadata that relates
to the sequencing run and are, for example, used by the EPI2ME software for the
preparation of the rich real-time reports. The `fishy_fastq` method used in the
previous section can also parse these sequence metadata facets from the FASTQ
file.

### Let's review the metadata content from a Guppy base-called FASTQ file


### Prepare a `tibble` from these Guppy based FASTQ data


