---
title: "BAM QC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BAM QC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows the curated BAM QC surface in floundeR. BAM parsing and
inspection belong to Bamana-backed Rust library code called in process from R;
floundeR keeps the R API focused on QC, review, provenance, reporting, and
Synoptikon handoff.

No BAM file is bundled with this vignette. The concrete BAM inspection calls are
therefore shown as a real workflow to run against local alignments, while the
offline report-card schema example remains evaluable without external data.

```{r setup}
library(floundeR)
```

## Read-Only BAM Evidence

For an aligned nanopore run, start with read-only evidence calls. These functions
return R-native lists and data frames while preserving Bamana's distinctions
between summary, verification, validation, EOF evidence, index state, mapping
state, sort evidence, and aux-tag evidence.

```{r bam-workflow, eval = FALSE}
bam_path <- "/path/to/alignments.bam"

bam_summary_evidence <- bam_summary(bam_path, include_mapq_hist = TRUE)
bam_verify_evidence <- bam_verify(bam_path)
bam_validation_evidence <- bam_validate(bam_path)
bam_eof_evidence <- bam_check_eof(bam_path)
bam_index_evidence <- bam_check_index(bam_path)
bam_mapping_evidence <- bam_check_map(bam_path)
bam_sort_evidence <- bam_check_sort(bam_path)
bam_rg_tag_evidence <- bam_check_tag(bam_path, "RG", full_scan = TRUE)
```

Transformation operations such as filtering, subsampling, sorting, merging,
unmapping, reheadering, or annotation are outside this read-only vignette. When
they are exposed through floundeR, they should require explicit output paths and
provenance reporting.

## BAM Report Card

`bam_qc_report_card()` consumes Bamana-derived evidence and evaluates it against
stable pass/warn/fail checks. It does not call Bamana directly, which keeps the
report-card contract testable and reusable by Synoptikon and Grammateus
reporting code.

```{r bam-report-card-workflow, eval = FALSE}
bam_card <- bam_qc_report_card(
  summary = bam_summary_evidence,
  index = bam_index_evidence,
  mapping = bam_mapping_evidence,
  sorting = bam_sort_evidence,
  tags = bam_rg_tag_evidence,
  eof = bam_eof_evidence,
  validation = bam_validation_evidence,
  provenance_anomalies = 0,
  expected_tags = c("RG")
)

bam_card
```

Without a local BAM file, the same function still shows the complete report-card
schema and marks absent evidence explicitly.

```{r bam-report-card-offline}
bam_card_schema <- bam_qc_report_card(expected_tags = c("RG"))
bam_card_schema[, c("check_id", "status", "observed_value", "details")]
```

## QC Handoff Shape

BAM evidence should be handed forward as sectioned QC tables rather than as
opaque log text. The report card can be attached to a Synoptikon payload or
turned into semantic Grammateus report elements when the optional private runtime
is available.

```{r bam-synoptikon-shape}
bam_payload <- as_synoptikon_qc(
  bam = list(report_card = bam_card_schema),
  report_cards = list(bam = bam_card_schema),
  input_provenance = list(
    input_id = "example-bam-qc-schema",
    kind = "bam",
    uri = "not-bundled",
    size_bytes = NA_real_,
    sha256 = NULL,
    metadata = list(package_vignette = TRUE)
  ),
  run_identity = list(
    run_id = "example-run",
    sample_id = "example-sample",
    project_id = "flounder-revival"
  ),
  payload_id = "example-bam-qc-payload",
  generated_at_utc = "2026-06-14T14:42:10Z"
)

names(bam_payload$qc_sections)
bam_payload$handoff
```

For an end-to-end handoff that combines run summary, POD5 provenance, BAM
evidence, and library-preparation evidence, see the Synoptikon handoff vignette.
