---
title: "fastq_processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastq_processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Preamble

This `fastq_processing` vignette re-introduces some of the content from the
`BasicQC` tutorial concept but derives the primary information from the FASTQ
file for users who have not maintained their sequencing_summary file. This is
intended to facilitate the development of workflows and reports that are
decoupled from the requirement for the `sequencing_summary` file.


## Getting started with the canned dataset

```{r setup}
library(floundeR)
```



## Exploring a more impressive sequence collection

The `Guppy` basecalling software converts the Nanopore format FAST5 raw sequence
files into the FASTQ files that we have reviewed in the previous section. The
FASTQ entries prepared by `Guppy` contain additional information in their
**header** fields. These additional information contain metadata that relates
to the sequencing run and are, for example, used by the EPI2ME software for the
preparation of the rich real-time reports. The `fishy_fastq` method used in the
previous section can also parse these sequence metadata facets from the FASTQ
file.

### Let's review the metadata content from a Guppy base-called FASTQ file


### Prepare a `tibble` from these Guppy based FASTQ data
