---
title: "Grammateus Technical Report"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grammateus Technical Report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how floundeR prepares a Grammateus-backed technical QC
report. The package vignette is rendered with RMarkdown, but the report artifact
it creates is not an RMarkdown report: it is a Grammateus semantic report
contract plus a manifest, provenance, figure metadata, and Mnemosyne Biosciences
theme references.

The public open-source package can build these contracts without private
Grammateus runtime assets. Governed HTML/PDF rendering is enabled only when an
authorized prebuilt Grammateus runtime is installed and validated.

```{r setup}
library(floundeR)
```

## Prepare QC Evidence

Start with run-level QC tables that are stable R objects and can also feed
Synoptikon handoff.

```{r qc-evidence}
summary_file <- flnDr("sequencing_summary.txt.bz2")
generated_at <- "2026-06-14T15:27:10Z"
run_id <- "example-grammateus-report"

run_summary <- qc_run_summary(
  summary_file,
  source_id = "example-sequencing-summary"
)
yield_over_time <- qc_yield_over_time(summary_file, resolution_minutes = 15)
quality_distribution <- qc_quality_distribution(summary_file, bins = 5)
flowcell_density <- qc_channel_density(summary_file)
barcode_balance <- qc_barcode_composition(summary_file)
run_report_card <- qc_report_card(
  run_summary,
  barcode_composition = barcode_balance
)

run_summary
run_report_card[, c("check_id", "status", "observed_value", "details")]
```

POD5, BAM, and library-preparation evidence remain sectioned report inputs. This
small vignette records POD5 open-data provenance and marks missing BAM/library
evidence explicitly rather than inventing bundled files.

```{r cross-surface-evidence}
pod5_example <- ont_zymo_pod5_example_objects(role = "pass")

pod5_integrity <- data.frame(
  schema_version = "flounder.pod5_open_data_example.v1",
  status = "not_downloaded",
  role = pod5_example$role,
  s3_uri = pod5_example$s3_uri,
  size_bytes = pod5_example$size,
  intended_use = pod5_example$intended_use,
  stringsAsFactors = FALSE
)

bam_alignment_summary <- data.frame(
  schema_version = "flounder.bam_alignment_summary.v1",
  status = "not_checked",
  detail = "No BAM file is bundled with this vignette fixture.",
  stringsAsFactors = FALSE
)

library_preparation <- data.frame(
  schema_version = "flounder.library_preparation_qc.v1",
  status = "not_checked",
  detail = paste(
    "No read-level library-preparation screen is bundled with this vignette.",
    "Porkchop-derived adapter, primer, barcode, kit, and cDNA evidence should",
    "populate this table when available."
  ),
  stringsAsFactors = FALSE
)
```

## Create A Governed Figure Wrapper

Figures handed to Grammateus need stable identifiers, captions, alt text,
artifact checksums, and provenance. This example wraps a tiny SVG fixture so the
vignette remains dependency-light; the R-to-Grammateus plot handoff vignette
covers `ggplot2` figure generation in more detail.

```{r report-figure}
report_dir <- file.path(tempdir(), "flounder-grammateus-report")
unlink(report_dir, recursive = TRUE)
dir.create(report_dir, recursive = TRUE)

yield_svg <- file.path(report_dir, "yield-over-time.svg")
writeLines(
  c(
    "<svg xmlns=\"http://www.w3.org/2000/svg\" width=\"240\" height=\"120\">",
    "  <rect width=\"240\" height=\"120\" fill=\"#ffffff\"/>",
    "  <polyline points=\"20,95 75,72 130,48 185,34 220,24\"",
    "    fill=\"none\" stroke=\"#237A57\" stroke-width=\"5\"/>",
    "  <circle cx=\"220\" cy=\"24\" r=\"6\" fill=\"#237A57\"/>",
    "</svg>"
  ),
  yield_svg,
  useBytes = TRUE
)

yield_figure <- grammateus_figure_from_file(
  path = yield_svg,
  figure_id = "figure_yield_over_time",
  caption = "Cumulative sequencing yield over time.",
  alt_text = paste(
    "Line chart icon showing cumulative sequencing yield increasing over",
    "elapsed run time."
  ),
  methods_note = "Wrapped as a deterministic SVG artifact for vignette use.",
  width_px = 240,
  height_px = 120,
  produced_at_utc = generated_at,
  run_id = run_id
)

yield_figure$source[c("format", "checksum", "width_px", "height_px")]
```

## Assemble Semantic Report Elements

`grammateus_qc_report_elements()` maps floundeR QC tables into named semantic
technical-report elements. The Mnemosyne theme descriptor records the private
runtime template/theme references without bundling private assets.

```{r report-elements}
run_metadata <- data.frame(
  run_id = run_id,
  sample_id = "example-sample",
  project_id = "flounder-revival",
  pod5_source = pod5_example$s3_uri,
  stringsAsFactors = FALSE
)

limitations <- data.frame(
  limitation_id = "example_pod5_not_downloaded",
  severity = "info",
  section = "pod5",
  message = "The selected ONT POD5 object is recorded as provenance only.",
  stringsAsFactors = FALSE
)

input_provenance <- data.frame(
  input_id = "example-sequencing-summary",
  kind = "sequencing_summary",
  uri = summary_file,
  size_bytes = unname(file.info(summary_file)$size),
  stringsAsFactors = FALSE
)

elements <- grammateus_qc_report_elements(
  run_metadata = run_metadata,
  qc_summary = run_summary,
  flowcell_density = flowcell_density,
  yield_over_time = yield_over_time,
  quality_distribution = quality_distribution,
  barcode_balance = barcode_balance,
  pod5_integrity = pod5_integrity,
  bam_alignment_summary = bam_alignment_summary,
  library_preparation = library_preparation,
  report_card_findings = run_report_card,
  methods = paste(
    "Run-level metrics were generated from a floundeR sequencing-summary",
    "fixture. POD5 source metadata references the canonical ONT Zymo fecal",
    "pass example without downloading raw POD5 bytes."
  ),
  limitations = limitations,
  provenance = input_provenance,
  produced_at_utc = generated_at,
  run_id = run_id
)

theme <- grammateus_mnemosyne_theme(
  produced_at_utc = generated_at,
  run_id = run_id
)

elements$element_count
theme$brand$name
theme$template$runtime_template_ref
```

## Write The Report Contract

`qc_report()` writes a Grammateus-shaped contract and manifest. The example uses
`render = "never"` so public vignette builds never require private runtime
assets. Use `render = "if_available"` or `render = "require"` in authorized
runtime environments.

```{r write-report}
report <- qc_report(
  elements = elements,
  figures = list(yield_figure),
  output_dir = report_dir,
  output = c("html", "pdf"),
  render = "never",
  theme = theme,
  report_id = "report_example_nanopore_qc",
  title = "Example nanopore sequencing QC report",
  produced_at_utc = generated_at,
  run_id = run_id
)

basename(report$contract_path)
basename(report$manifest_path)
report$manifest[c("schema_version", "report_id", "element_count", "figure_count")]
vapply(report$outputs, `[[`, character(1), "status")
```

The manifest records the contract checksum, output status, theme references, and
provenance needed for governed report lifecycle handling.

```{r inspect-manifest}
manifest <- jsonlite::fromJSON(report$manifest_path, simplifyVector = FALSE)

manifest$contract[c("sha256", "byte_len")]
manifest$theme
manifest$runtime$status
```

## Runtime-Backed Rendering

Authorized environments can request runtime-backed HTML/PDF rendering after the
prebuilt Grammateus runtime has been installed and validated.

```{r runtime-rendering, eval = FALSE}
validation <- grammateus_runtime_validate()

if (isTRUE(validation$ok)) {
  rendered <- qc_report(
    elements = elements,
    figures = list(yield_figure),
    output_dir = report_dir,
    output = c("html", "pdf"),
    render = "if_available",
    report_id = "report_example_nanopore_qc",
    title = "Example nanopore sequencing QC report",
    produced_at_utc = generated_at,
    run_id = run_id
  )
}
```

Use `render = "require"` only when a missing or invalid private runtime should
be treated as a hard failure.
