---
title: "Grammateus Plot Directions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grammateus Plot Directions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows the two supported plot directions in floundeR's Grammateus
reporting interface:

- an analyst-owned R `ggplot2` plot is dropped into a Grammateus report contract;
- a Grammateus semantic plot specification is rendered by controlled R code and
  then included in a report contract.

The vignette itself is RMarkdown documentation. The report artifacts it creates
are Grammateus-shaped semantic contracts and manifests, and governed HTML/PDF
rendering remains optional through the private prebuilt Grammateus runtime.

```{r setup}
library(floundeR)
```

## Prepare Shared QC Evidence

Both directions start with ordinary floundeR QC tables. They are stable R
objects, so they can be inspected interactively, exported to Synoptikon, or
assembled into Grammateus report elements.

```{r qc-evidence}
summary_file <- flnDr("sequencing_summary.txt.bz2")
generated_at <- "2026-06-14T15:42:10Z"
run_id <- "example-grammateus-plot-directions"

report_dir <- file.path(tempdir(), "flounder-grammateus-plot-directions")
unlink(report_dir, recursive = TRUE)
dir.create(report_dir, recursive = TRUE)

run_summary <- qc_run_summary(
  summary_file,
  source_id = "example-sequencing-summary"
)
yield_over_time <- qc_yield_over_time(summary_file, resolution_minutes = 15)
run_report_card <- qc_report_card(run_summary)

elements <- grammateus_qc_report_elements(
  qc_summary = run_summary,
  yield_over_time = yield_over_time,
  report_card_findings = run_report_card,
  methods = paste(
    "Run-level QC was generated from the bundled sequencing-summary fixture.",
    "Plot artifacts were produced through the floundeR Grammateus reporting",
    "interface without requiring private runtime assets."
  ),
  provenance = data.frame(
    input_id = "example-sequencing-summary",
    kind = "sequencing_summary",
    uri = summary_file,
    size_bytes = unname(file.info(summary_file)$size),
    stringsAsFactors = FALSE
  ),
  produced_at_utc = generated_at,
  run_id = run_id
)

theme <- grammateus_mnemosyne_theme(
  produced_at_utc = generated_at,
  run_id = run_id
)

run_summary
```

## Direction 1: R Plot Dropped Into A Report

Use this path when an analyst has already produced a `ggplot2` plot in R and
needs to hand it to the governed report pipeline with stable identifiers,
captions, alt text, artifact checksums, and source-data provenance.

```{r r-to-grammateus}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  yield_plot <- ggplot2::ggplot(
    yield_over_time,
    ggplot2::aes(
      x = bin_end_minutes,
      y = cumulative_bases,
      colour = passes_filtering,
      group = passes_filtering
    )
  ) +
    ggplot2::geom_line(linewidth = 0.7) +
    ggplot2::labs(
      x = "Elapsed time (minutes)",
      y = "Cumulative bases",
      colour = "Passes filtering"
    ) +
    ggplot2::theme_minimal(base_size = 11)

  r_owned_figure <- grammateus_figure_from_ggplot(
    plot = yield_plot,
    figure_id = "figure_r_owned_yield_over_time",
    caption = "Cumulative sequencing yield over time from an analyst R plot.",
    alt_text = paste(
      "Line plot showing cumulative sequencing bases over elapsed run time,",
      "split by whether reads passed filtering."
    ),
    output_dir = report_dir,
    formats = "png",
    width = 140,
    height = 90,
    units = "mm",
    dpi = 120,
    methods_note = "Generated in R with ggplot2 and wrapped for Grammateus.",
    source_data = yield_over_time,
    produced_at_utc = generated_at,
    run_id = run_id
  )

  r_owned_figure$source[c("format", "path", "checksum")]
} else {
  r_owned_figure <- NULL
  data.frame(
    status = "not_checked",
    reason = "ggplot2 is not installed in this vignette environment."
  )
}
```

The figure can be inserted into a report contract without private runtime
assets. `render = "never"` deliberately writes the semantic contract and
manifest only; authorized environments can switch to `render = "if_available"`
or `render = "require"` after validating a private Grammateus runtime.

```{r r-owned-report}
if (!is.null(r_owned_figure)) {
  r_owned_report <- qc_report(
    elements = elements,
    figures = list(r_owned_figure),
    output_dir = file.path(report_dir, "r-owned-report"),
    output = c("html", "pdf"),
    render = "never",
    theme = theme,
    report_id = "report_r_owned_plot_handoff",
    title = "Example QC report with an R-owned plot",
    produced_at_utc = generated_at,
    run_id = run_id
  )

  r_owned_report$manifest[c(
    "schema_version",
    "report_id",
    "element_count",
    "figure_count"
  )]
}
```

## Direction 2: Grammateus Plot Spec Rendered By R

Use this path when a report template, governed workflow, or Synoptikon-facing
process owns the plot definition. floundeR builds a backend-neutral semantic
plot specification from tidy QC data, then the controlled R backend renders the
requested image artifacts in a deterministic run directory.

```{r grammateus-to-r}
plot_spec <- qc_plot_yield_over_time(
  yield_over_time,
  output = list(
    width_mm = 140,
    height_mm = 90,
    dpi = 120,
    formats = "png"
  ),
  produced_at_utc = generated_at,
  run_id = run_id
)

plot_spec[c("schema_version", "plot_id", "plot_type", "caption")]
plot_spec$mappings
```

The render step needs local R plot dependencies and an `Rscript` executable. It
does not need the private Grammateus HTML/PDF runtime because this is the
controlled R plot backend, not final report rendering.

```{r render-plot-spec}
can_render_plot <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("jsonlite", quietly = TRUE) &&
  requireNamespace("scales", quietly = TRUE) &&
  requireNamespace("viridisLite", quietly = TRUE) &&
  nzchar(Sys.which("Rscript"))

if (can_render_plot) {
  plot_run <- grammateus_render_plot(
    plot_spec,
    execution = "local_rscript",
    run_root = file.path(report_dir, "plot-runs"),
    required_packages = c("ggplot2", "jsonlite", "scales", "viridisLite")
  )

  spec_rendered_figure <- plot_run$artifacts$png$figure

  list(
    run_dir = plot_run$run_dir,
    script_sha256 = plot_run$script_sha256,
    figure_checksum = spec_rendered_figure$source$checksum
  )
} else {
  plot_run <- NULL
  spec_rendered_figure <- NULL
  data.frame(
    status = "not_checked",
    reason = paste(
      "The controlled R plotting backend needs ggplot2, jsonlite, scales,",
      "viridisLite, and Rscript."
    )
  )
}
```

The rendered figure uses the same `flounder_grammateus_figure` wrapper as an
analyst-supplied R plot, so downstream report assembly does not need a second
figure contract.

```{r spec-owned-report}
if (!is.null(spec_rendered_figure)) {
  spec_owned_report <- qc_report(
    elements = elements,
    figures = list(spec_rendered_figure),
    output_dir = file.path(report_dir, "spec-owned-report"),
    output = "html",
    render = "never",
    theme = theme,
    report_id = "report_spec_owned_plot_render",
    title = "Example QC report with a Grammateus-owned plot spec",
    produced_at_utc = generated_at,
    run_id = run_id
  )

  spec_owned_report$manifest[c(
    "schema_version",
    "report_id",
    "element_count",
    "figure_count"
  )]
}
```

## Practical Boundary

The two plot directions are deliberately symmetrical at report assembly time:
both produce semantic figure metadata with stable identifiers, captions, alt
text, checksums, source-data hashes, and run provenance. The difference is who
owns the plot definition:

- R-owned plots are best for interactive analyst work and specialized
  exploratory figures.
- Grammateus-owned plot specs are best for reproducible technical reports,
  governed templates, and Synoptikon report lifecycle handoff.

Neither path downloads data, shells out to a Grammateus CLI, or requires private
Grammateus source code during public package checks.
