---
title: "POD5 QC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{POD5 QC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows the package-level POD5 QC workflow in floundeR. It is a
package tutorial, not a production report renderer. Governed HTML/PDF report
artifacts remain a Grammateus responsibility.

The default workflow is network-free and does not download or commit POD5 bytes.
Rust-backed POD5 inspection is demonstrated only when the compiled extension is
available in the local R installation.

```{r setup}
library(floundeR)
rust_available <- flounder_rust_available()
```

## Canonical Open-Data Source

floundeR uses a fixed ONT open-data POD5 source for opt-in real-data examples:
`s3://ont-open-data/zymo_fecal_2025.05/raw/PAU85136/pod5/`.

The routine pass example is
`PAU85136_pass_279c9095_68316534_8289.pod5`. The fail object is reserved for
fail-state examples and should not be used as the default tutorial input.

```{r open-data-metadata}
pod5_dataset <- ont_zymo_pod5_dataset()
pod5_examples <- ont_zymo_pod5_example_objects()

pod5_dataset
pod5_examples
```

Opt in explicitly before downloading from S3, and keep cached files outside the
repository:

```{r opt-in-fetch, eval = FALSE}
pass_example <- ont_zymo_pod5_example_objects(role = "pass")

if (identical(tolower(Sys.getenv("FLOUNDER_RUN_NETWORK_TESTS")), "true")) {
  fetched <- ont_open_data_fetch(
    key = pass_example$key,
    cache_dir = tools::R_user_dir("floundeR", which = "cache")
  )

  pod5_file_info(fetched$cache_path)
  pod5_verify(fetched$cache_path)
}
```

## Local POD5 Inventory

The curated R API is intentionally small: discover POD5-containing folders,
verify candidate files, inspect file or folder metadata, build manifests, compare
collections, and plan read-only demonstration subdivisions. These calls go
through in-process Rust and return R data frames or lists.

The tiny files below are POD5-like signature fixtures for documentation. They
exercise discovery, manifest, and envelope checks without embedding real raw
signal data in the package.

```{r local-pod5-inventory}
if (rust_available) {
  signature <- as.raw(c(0x8b, 0x50, 0x4f, 0x44, 0x0d, 0x0a, 0x1a, 0x0a))
  run_dir <- file.path(tempdir(), "flounder-pod5-qc-example")
  unlink(run_dir, recursive = TRUE)
  dir.create(file.path(run_dir, "pass"), recursive = TRUE)
  dir.create(file.path(run_dir, "fail"), recursive = TRUE)

  writeBin(
    c(signature, as.raw(rep(0, 16)), signature),
    file.path(run_dir, "pass", "example-pass.pod5")
  )
  writeBin(
    c(signature, as.raw(rep(1, 16)), signature),
    file.path(run_dir, "fail", "example-fail.pod5")
  )

  pod5_find(run_dir)
  pod5_folder_info(run_dir)
  pod5_manifest(run_dir)
} else {
  data.frame(
    status = "not_checked",
    reason = "Compiled Rust support is unavailable in this R installation.",
    stringsAsFactors = FALSE
  )
}
```

File-level inspection separates fast envelope evidence from unavailable deep
POD5 metadata. That prevents a lightweight tutorial fixture from being
overstated as fully parsed raw-signal evidence.

```{r local-pod5-file}
if (rust_available) {
  example_file <- file.path(run_dir, "pass", "example-pass.pod5")

  pod5_verify(example_file)
  pod5_file_info(example_file)
} else {
  data.frame(
    status = "not_checked",
    reason = "Compiled Rust support is unavailable in this R installation.",
    stringsAsFactors = FALSE
  )
}
```

## Demonstration Subdivision Plans

Large POD5 files are not suitable for routine vignette builds. When a real POD5
object has been cached outside the repository, floundeR can ask the Rust-backed
POD5 layer for a read-only split plan. This plans demonstration subsets without
writing new POD5 files.

```{r pod5-subdivide-plan}
if (rust_available) {
  pod5_subdivide_plan(run_dir, strategy = "file-count", files_per_chunk = 1)
} else {
  data.frame(
    status = "not_checked",
    reason = "Compiled Rust support is unavailable in this R installation.",
    stringsAsFactors = FALSE
  )
}
```

Actual subset or split file creation remains an explicit derived-data workflow:
record the source URI, checksum, command/runtime provenance, and generated
manifest, then keep the derived POD5 artifact outside the package repository.

## QC Handoff Shape

POD5 evidence is normally passed forward as manifest, verification, and
open-data provenance tables. Those tables can be attached to Synoptikon handoff
payloads or converted into Grammateus report elements when the optional private
runtime is installed.

```{r pod5-handoff-shape}
pass_example <- ont_zymo_pod5_example_objects(role = "pass")

pod5_provenance <- data.frame(
  schema_version = "flounder.pod5_open_data_example.v1",
  status = "not_downloaded",
  bucket = pass_example$bucket,
  region = pass_example$region,
  key = pass_example$key,
  s3_uri = pass_example$s3_uri,
  size_bytes = pass_example$size,
  intended_use = pass_example$intended_use,
  stringsAsFactors = FALSE
)

pod5_provenance
```

Use the companion Synoptikon handoff vignette for the full JSON payload contract.
