---
title: "sequencing_summary_review"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sequencing_summary_review}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Preamble

This `sequencing_summary_review` vignette re-introduces the `BasicQC`-type
tutorial that was previously available through the https://github.com/nanoporetech
website.

The workflow provided is implemented through the R TidyVerse and is intended to
provide a lightweight set of components that can be mixed and matched for the
exploration of the sequencing_summary file that is produced during Guppy-based
base-calling.

```{r setup}
library(floundeR)
```


## Define input files and create S4 object

```{r create_ss}
sequencing_summary <- system.file("extdata",
                                  "sequencing_summary.txt.bz2", 
                                  package="floundeR")
barcodes_summary <- system.file("extdata",
                                "barcoding_summary.txt.bz2", 
                                package="floundeR")

```
