---
title: "Synoptikon QC Handoff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Synoptikon QC Handoff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows the package-level handoff contract from floundeR QC data to
Synoptikon/Mneion. It is an R package vignette, not the production QC report
renderer. Governed HTML/PDF report rendering remains a Grammateus responsibility
and is intentionally optional for the open-source floundeR package.

The example is network-free. It uses the package sequencing-summary fixture and
metadata for the selected ONT open-data POD5 object without downloading POD5
bytes.

For focused input-level tutorials, see the POD5 QC and BAM QC vignettes. This
vignette combines those evidence shapes into the Synoptikon JSON handoff
contract.

```{r setup}
library(floundeR)
```

## Build Run-Level QC Evidence

Start from a sequencing-summary file and derive the stable QC tables that
Synoptikon can ingest directly.

```{r run-qc}
summary_file <- flnDr("sequencing_summary.txt.bz2")

run_summary <- qc_run_summary(
  summary_file,
  source_id = "example-sequencing-summary"
)

flowcell_density <- qc_channel_density(summary_file)
barcode_composition <- qc_barcode_composition(summary_file)
run_report_card <- qc_report_card(summary_file)

run_summary
run_report_card
```

## Add POD5 Source Metadata

The canonical contemporary POD5 source for floundeR demonstrations is the ONT
Zymo fecal run at `s3://ont-open-data/zymo_fecal_2025.05/raw/PAU85136/pod5/`.
The routine pass example is the smallest pass POD5 object recorded for that
prefix. This vignette records the selected object as provenance without
downloading it.

```{r pod5-metadata}
pod5_dataset <- ont_zymo_pod5_dataset()
pod5_example <- ont_zymo_pod5_example_objects(role = "pass")

pod5_handoff <- data.frame(
  schema_version = "flounder.pod5_open_data_example.v1",
  status = "not_checked",
  bucket = pod5_example$bucket,
  region = pod5_example$region,
  key = pod5_example$key,
  s3_uri = pod5_example$s3_uri,
  size_bytes = pod5_example$size,
  last_modified_utc = pod5_example$last_modified_utc,
  intended_use = pod5_example$intended_use,
  stringsAsFactors = FALSE
)

pod5_dataset
pod5_handoff
```

To fetch and inspect the selected POD5 object, opt in explicitly and cache the
file outside the repository:

```{r pod5-fetch, eval = FALSE}
if (identical(tolower(Sys.getenv("FLOUNDER_RUN_NETWORK_TESTS")), "true")) {
  fetched <- ont_open_data_fetch(
    key = pod5_example$key,
    cache_dir = tools::R_user_dir("floundeR", which = "cache")
  )

  pod5_info <- pod5_file_info(fetched$cache_path)
  pod5_integrity <- pod5_verify(fetched$cache_path)
}
```

## Represent Reserved Evidence Surfaces

BAM/Bamana and Porkchop/library-preparation evidence are first-class handoff
sections. In this minimal vignette they are intentionally marked as not checked,
because no BAM file or library-preparation screen is bundled with the package
fixture.

```{r reserved-evidence}
bam_evidence <- data.frame(
  schema_version = "flounder.bam_qc_report_card.v1",
  status = "not_checked",
  check_id = "bam_evidence_not_supplied",
  detail = "No BAM file is bundled with this vignette fixture.",
  stringsAsFactors = FALSE
)

library_preparation_evidence <- data.frame(
  schema_version = "flounder.library_preparation_qc.v1",
  status = "not_checked",
  evidence_kind = "porkchop_reserved",
  detail = paste(
    "Porkchop adapter, primer, barcode, kit, and cDNA evidence can populate",
    "this section when library-preparation screening evidence is supplied."
  ),
  stringsAsFactors = FALSE
)
```

## Create The Synoptikon Payload

`as_synoptikon_qc()` assembles the sectioned evidence in memory.
`write_synoptikon_qc()` serialises the same contract as JSON for handoff.

```{r payload}
payload <- as_synoptikon_qc(
  run_summary = run_summary,
  report_cards = run_report_card,
  flowcell = list(channel_density = flowcell_density),
  barcode = list(composition = barcode_composition),
  pod5 = list(open_data_example = pod5_handoff),
  bam = list(report_card = bam_evidence),
  library_preparation = list(evidence = library_preparation_evidence),
  input_provenance = list(
    input_id = "example-sequencing-summary",
    kind = "sequencing_summary",
    uri = summary_file,
    size_bytes = unname(file.info(summary_file)$size),
    sha256 = NULL,
    metadata = list(package_fixture = TRUE)
  ),
  run_identity = list(
    run_id = "example-run",
    sample_id = "example-sample",
    project_id = "flounder-revival",
    flow_cell_id = NA_character_
  ),
  governance_context = list(
    tenant_id = "example-tenant",
    governance_domain_id = "example-domain",
    mneion_work_request_id = "example-work-request"
  ),
  limitations = list(
    limitation_id = "example-pod5-not-downloaded",
    severity = "info",
    section = "pod5",
    message = paste(
      "This vignette records ONT open-data POD5 provenance but does not",
      "download POD5 bytes by default."
    )
  ),
  payload_id = "example-synoptikon-qc-payload",
  generated_at_utc = "2026-06-14T06:39:36Z"
)

names(payload)
names(payload$qc_sections)
payload$handoff
```

```{r write-json}
handoff_path <- file.path(tempdir(), "flounder-synoptikon-qc-payload.json")
write_synoptikon_qc(
  handoff_path,
  run_summary = run_summary,
  report_cards = run_report_card,
  flowcell = list(channel_density = flowcell_density),
  barcode = list(composition = barcode_composition),
  pod5 = list(open_data_example = pod5_handoff),
  bam = list(report_card = bam_evidence),
  library_preparation = list(evidence = library_preparation_evidence),
  input_provenance = payload$input_provenance,
  run_identity = payload$run_identity,
  governance_context = payload$governance_context,
  limitations = payload$limitations,
  payload_id = payload$payload_id,
  generated_at_utc = payload$generated_at_utc
)

jsonlite::fromJSON(handoff_path, simplifyVector = FALSE)$schema_version
```

The written JSON follows the installed schema at
`inst/schema/synoptikon-qc-payload-v1.schema.json`. Downstream Synoptikon
ingestion can use that schema to validate the handoff before creating governed
workbench records, report-build inputs, or audit evidence.

## Where Grammateus Fits

The Synoptikon payload is QC evidence. It is not a trusted report manifest and
does not replace Grammateus lifecycle contracts. Once Grammateus runtime support
lands, the same sectioned QC data should feed semantic report elements, governed
figures, provenance tables, and Mnemosyne Biosciences branded HTML/PDF reports.
